A Neurally-Guided, Parallel Theorem Prover

### 2019-07-01 papers/FroCoS19/paper.tex
final FroCoS paper

\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neurally-Guided, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
}
\maketitle

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
However, in practice there are several implementation issues~\cite{lerna-discussion} which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.

@@

%A formula is either true \(\top\), false \(\bot\), an equality between terms \(t_1 = t_2\), a predicate applied to terms \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol, a negated formula \(\lnot \phi\), an impliciation \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
%Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
%Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).

\paragraph{Automatic Theorem Provers}
An automated theorem prover (ATP) is a system able to automatically establish whether a formula (in first-order logic) is \emph{satisfiable} or \emph{unsatisfiable}; although, given the undecidability of this problem, ATPs may also return \emph{unknown}. Both saturation-based provers (e.g. E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}) and SMT solvers (e.g. CVC4~\cite{CVC4} and Z3~\cite{Z3}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
%Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
%Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.

@@

Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode the syntactic structure of proof state and therefore lose information.
By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.

@@

\paragraph{The MPTP Problem Set}
For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).

>>>>

\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neurally-Guided, Parallel Theorem Prover}
\author{
	Michael Rawson\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
}
\maketitle

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
However, in practice there are several implementation issues~\cite{lerna-discussion} which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
\begin{enumerate}
	\item Proof state in such systems may be of impractical size, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate. In a saturation context, the size of the current proof state may be many times the size of the eventual proof: while neural networks are in principle capable of processing large amounts of data, throughput suffers and scalability is a concern.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}. Source code available at \url{https://github.com/MichaelRawson/lerna}.} takes an alternative step toward useful neural automatic theorem provers.

@@

\paragraph{Automatic Theorem Provers}
An automated theorem prover (ATP) is a system able to automatically establish whether a formula (in first-order logic) is \emph{satisfiable} or \emph{unsatisfiable}; although, given the undecidability of this problem, ATPs may also return \emph{unknown}. Both saturation-based provers (e.g. E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}) and SMT solvers (e.g. CVC4~\cite{CVC4} and Z3~\cite{Z3}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.

@@

Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA,proof-mining}, which have the advantage of simplicity and relative efficiency, but do not fully encode the syntactic structure of proof state and therefore lose information.
By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.

@@

\paragraph{The MPTP Problem Set}
For training and evaluation purposes a set of valid propositions exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).
